# Explanatory Data Analysis

## Packages

```{r}
#| warning: false
library(dplyr)
library(knitr)
```

## Dataset

Dataset is a synthetic version of the original dataset due to disclosure control. Synthetic version was generated with `synthesizer` package available on [CRAN](https://CRAN.R-project.org/package=synthesizer).

First, we load the dataset.

```{r}
#| warning: false
#| eval: true
url <- "https://github.com/AIML4OS/WP9_S1_PL_TOURISM/raw/refs/heads/main/sspcloud/WP9_S1_Tourism_synt.rds"

# Temporary file
temp_file <- tempfile(fileext = ".rds")

# Download data
download.file(url, temp_file, mode = "wb")

# Read RDS
data <- readRDS(temp_file)
```

```{r}
#| echo: false
#| eval: true
data$seasonality=NULL
```


Check structure of the dataset

```{r}
#| warning: false
#| eval: true
head(data) %>% kable()
```

These variables are interpreted as follows:

- `regon` - ID of accommodation establishment (original Polish name of ID but content is completely different)
- `voivodeship` - administrative region code (closely corresponds to NUTS 2)
- `county` - administrative region code (former LAU 2)
- `municipality` - administrative region code (current LAU, former LAU 1)
- `object_type` - recoded type of accommodation establishments, e.g. hotel, motel, etc.
- `is_seasonal` - `1` when operates through whole year, `2` - when operates in a part of a year (e.g. in a summer or in a winter season)
- `period_yyyymm` - text label of year and month of reporting (convenient for data ordering)
- `tourists` - number of tourists accommodated (variable of interest)
- `month_num` - number of month (`1` for January)
- `active_in_month` - `1` when establishment operated in a given month (there are many cases when establishment does not provide number of tourists but declares it was active)
- `year` - year as integer
- `month` - text label of the month


# Conclusion
