# Explanatory Data Analysis

## Packages

```{r}
#| warning: false
#| message: false
library(dplyr)
library(knitr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(zoo)
library(corrplot)
```

## Dataset

Dataset is a synthetic version of the original dataset due to disclosure control. Synthetic version was generated with `synthesizer` package available on [CRAN](https://CRAN.R-project.org/package=synthesizer).

First, we load the dataset.

```{r}
#| warning: false
#| eval: true
url <- "https://github.com/AIML4OS/WP9_S1_PL_TOURISM/raw/refs/heads/main/sspcloud/WP9_S1_Tourism_synt.rds"

# Temporary file
temp_file <- tempfile(fileext = ".rds")

# Download data
download.file(url, temp_file, mode = "wb")

# Read RDS
data <- readRDS(temp_file)
```

Check structure of the dataset

```{r}
#| warning: false
#| eval: true
head(data) %>% kable()
```

These variables are interpreted as follows:

- `regon` - ID of accommodation establishment (original Polish name of ID but content is completely different)
- `voivodeship` - administrative region code (closely corresponds to NUTS 2)
- `county` - administrative region code (former LAU 2)
- `municipality` - administrative region code (current LAU, former LAU 1)
- `object_type` - recoded type of accommodation establishments, e.g. hotel, motel, etc.
- `is_seasonal` - `1` when operates through whole year, `2` - when operates in a part of a year (e.g. in a summer or in a winter season)
- `period_yyyymm` - text label of year and month of reporting (convenient for data ordering)
- `tourists` - number of tourists accommodated (variable of interest)
- `month_num` - number of month (`1` for January)
- `active_in_month` - `1` when establishment operated in a given month (there are many cases when establishment does not provide number of tourists but declares it was active)
- `year` - year as integer
- `month` - text label of the month

Some additional variables will make furthere analysis more convinient


```{r}
#| echo: false
#| eval: true
data <- data %>%
  mutate(
    period_yyyymm = as.character(period_yyyymm),
    period_date = ym(period_yyyymm),        # lub lubridate::ym()
    year = as.integer(year),
    month_num = as.integer(month_num),
    month = sprintf("%02d", month_num),
    seasonality=if_else(is_seasonal==1,'year-round','seasonal')
  )
```



## Summary statistics

```{r}
#| warning: false
stats_table <- data %>%
  group_by(year, month, object_type) %>%
  summarise(
    min = min(tourists, na.rm = TRUE),
    q1 = quantile(tourists, 0.25, na.rm = TRUE),
    median = median(tourists, na.rm = TRUE),
    mean = mean(tourists, na.rm = TRUE),
    q3 = quantile(tourists, 0.75, na.rm = TRUE),
    max = max(tourists, na.rm = TRUE),
    na_count = sum(is.na(tourists)),
    .groups = "drop"
  )
```



Number of accommodation establishments, broken down into seasonal and year-round.

```{r}
#| warning: false
objects_summary <- data %>%
  group_by(year, month) %>%
  summarise(
    total_objects = n_distinct(regon),  # unique IDs
    seasonal_objects = sum(seasonality == "seasonal", na.rm = TRUE),
    year_round_objects = sum(seasonality == "year-round", na.rm = TRUE),
    .groups = "drop"
  )
```

Total number of tourists by accommodation establishments type.

```{r}
tourists_by_type <- data %>%
  group_by(object_type) %>%
  summarise(
    total_tourists = sum(tourists, na.rm = TRUE),
    .groups = "drop"
  )
```











#hist
```{r}
library(scales)
data %>% filter(tourists>0) %>% 
ggplot(aes(x = tourists)) +
  geom_histogram(bins = 80, fill = "#2C3E50", color = "white", alpha = 0.8) +
  scale_x_continuous(trans = "log10", labels = label_comma()) +  
  labs(
    title = "Histogram of Tourists (Logarithmic Scale)",
    x = "Number of Tourists (log10)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```



Seasonality of total number of turists
```{r}
tot_month <- data %>%
  group_by(period_date) %>%
  summarise(total_tourists = sum(tourists, na.rm=TRUE))

ggplot(tot_month, aes(x = period_date, y = total_tourists)) +
  geom_line() +
  labs(title = "Całkowita liczba turystów w czasie", x = "Miesiąc", y = "Liczba")
```

monthly_tourists <- data %>%
  group_by(period_date) %>%
  summarise(total_tourists = sum(tourists, na.rm = TRUE)) %>%
  arrange(period_date) %>%
  mutate(month_label = format(period_date, "%Y-%m"))

 #Wykres słupkowy z wartościami nad słupkami
sezo=ggplot(monthly_tourists, aes(x = month_label, y = total_tourists)) +
  geom_col(fill = "#1f77b4") +  # zmienny kolor słupków
  geom_text(aes(label = comma(total_tourists)),
            hjust = -0.5, size = 3.5, fontface = "bold", angle = 90) +  # przesunięcie nad słupkiem
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) + # więcej miejsca nad słupkami
  labs(
    title = "Monthly Tourist Volume",
    x = "Month",
    y = "Number of tourists"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16)
  )
ggsave("Monthly Tourist Volume.png", plot = sezo, width = 10, height = 6, dpi = 300)


season_plot <- data %>%
  group_by(month_num) %>%
  summarise(mean_t = mean(tourists, na.rm=TRUE),
            sd_t = sd(tourists, na.rm=TRUE))

ggplot(season_plot, aes(x = factor(month_num, levels = 1:12), y = mean_t)) +
  geom_line(aes(group=1)) +   # trzeba dodać group=1, żeby linia nie "znikła"
  geom_point() +
  labs(title = "Średnia liczba turystów wg miesiąca (agregat)",
       x = "Miesiąc", y = "Średnia") +
  scale_x_discrete(labels = 1:12)




# na typ obiektu
fac=ggplot(data %>% filter(!is.na(tourists)), aes(x = object_type, y = tourists)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Distribution of tourists by type of facility")
ggsave("Distribution of tourists by type of facility.png", plot = fac, width = 10, height = 6, dpi = 300)

# trend 
top_voiv <- data %>% 
  count(voivodeship, sort = TRUE) %>% 
  slice_head(n = 16) %>% 
  pull(voivodeship)

tre=ggplot(filter(data, voivodeship %in% top_voiv), aes(x = period_date, y = tourists, color = voivodeship, group = voivodeship)) +
  stat_summary(geom="line", fun = "mean")
ggsave("Average number of tourists over time in top 16 voivodeships_1.png", plot = tre, width = 10, height = 6, dpi = 300)

tre2=ggplot(data %>% filter(voivodeship %in% top_voiv), 
       aes(x = period_date, y = tourists, color = voivodeship, group = voivodeship)) +
  stat_summary(geom = "line", fun = mean, size = 1) +
  stat_summary(geom = "point", fun = mean, size = 2) +  # opcjonalnie, żeby widać punkty średnich
  labs(
    title = "Average number of tourists over time in top 16 voivodeships",
    x = "Date",
    y = "Average number of tourists",
    color = "Voivodeship"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("Average number of tourists over time in top 16 voivodeships_2.png", plot = tre2, width = 10, height = 6, dpi = 300)


Mapy

  
library(ggplot2)
library(sf)
library(dplyr)
library(RColorBrewer)
library(gridExtra)

# 0. Rozpakowanie pliku zip do folderu tymczasowego
zip_file <- "wojewodztwa.zip"  # Twój plik zip
temp_dir <- tempdir()           # tymczasowy folder
unzip(zip_file, exdir = temp_dir)

# 2. Znalezienie pliku .shp w rozpakowanym folderze
shp_file <- list.files(temp_dir, pattern = "\\.shp$", full.names = TRUE)

# 3. Wczytanie shapefile
poland_map <- st_read(shp_file)


# 4. Przygotowanie danych rocznych na województwo
# Załóżmy, że kolumna 'voivodeship' to numery województw
tur_year <- data %>%
  mutate(year = substr(period_yyyymm, 1, 4)) %>%
  group_by(voivodeship, year) %>%
  summarise(tourists = sum(tourists, na.rm = TRUE), .groups = "drop")

# 5. Zamiana numerów województw na nazwy (dopasuj do shapefile)
voivodeship_names <- c(
  "02"="Dolnośląskie", "04"="Kujawsko-Pomorskie", "06"="Lubelskie",
  "08"="Lubuskie", "10"="Łódzkie", "12"="Małopolskie",
  "14"="Mazowieckie", "16"="Opolskie", "18"="Podkarpackie",
  "20"="Podlaskie", "22"="Pomorskie", "24"="Śląskie",
  "26"="Świętokrzyskie", "28"="Warmińsko-Mazurskie", "30"="Wielkopolskie",
  "32"="Zachodniopomorskie"
)

tur_year <- tur_year %>%
  mutate(voivodeship_name = tolower(voivodeship_name))

# 6. Funkcja do rysowania mapy dla pojedynczego roku
draw_map <- function(df_year, rok){
  map_df <- poland_map %>%
    left_join(df_year, by = c("JPT_NAZWA_" = "voivodeship_name"))
  ggplot(map_df) +
    geom_sf(aes(fill = tourists), color = "white") +
    scale_fill_gradientn(colors = brewer.pal(9, "Blues"),
                         na.value = "grey90",
                         labels = scales::comma) +
    labs(title = paste("Year", rok), fill = "Number of tourists") +
    theme_minimal()
}

# 7. Przygotowanie map dla lat 2018-2020
map_2018 <- draw_map(filter(tur_year, year == "2018"), 2018)
map_2019 <- draw_map(filter(tur_year, year == "2019"), 2019)
map_2020 <- draw_map(filter(tur_year, year == "2020"), 2020)

# 8. Wyświetlenie map obok siebie
grid.arrange(map_2018, map_2019, map_2020, ncol = 3)

png("maps_2018_2020_grid.png", width = 2000, height = 800, res = 150)
grid.arrange(map_2018, map_2019, map_2020, ncol = 3)
dev.off()

# 2018
png("map_2018.png", width = 800, height = 800, res = 150)
print(map_2018)
dev.off()

# 2019
png("map_2019.png", width = 800, height = 800, res = 150)
print(map_2019)
dev.off()

# 2020
png("map_2020.png", width = 800, height = 800, res = 150)
print(map_2020)
dev.off()
