# Feature Engineering

Feature engineering is the process of transforming raw data into meaningful features that enhance the performance of machine learning models. In this chapter, the emphasis is put on feature creation and feature selection. Feature creation refers to deriving new variables using domain expertise or by detecting patterns within the data, while feature selection focuses on identifying and retaining the most relevant subset of features for analysis.

## Packages

```{r}
#| warning: false
library(dplyr)
library(knitr)
```

## Preliminaries

In this chapter we shall construct micro-level and mezo-evel variables with respect to time, spatial, and other dimensions. Optional data indicating the severity of restrictions during COVID will be included also. The following file contains monthly indicators of severity of COVID-19 restrictions.

```{r}
#| warning: false
#| eval: false
monthly_string=read.csv2("https://github.com/AIML4OS/WP9_S1_PL_TOURISM/raw/refs/heads/main/sspcloud/stringency_data.csv") #
```

Our data is already in long format, that is, every accommodation establishment is represented by several rows. Each row covers data for one month. 